Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates

### 2025-11-05 paper/1_Initial/manuscript.qmd
minor
incomplete conclusion

<!-- Aims -->
The present study provides a comprehensive evaluation of the IAS, examining both its structure (study 1) and its dispositional correlates (study 2). These studies offer novel insights into the conceptual interpretation, and practical applications of the IAS. In this discussion, these findings are contextualized within the broader landscape of interoception research.

<!-- Study 1 findings -->

In Study 1, the factor structure of the IAS was reanalysed, leading to a refined 14-item version of the scale. A shorter IAS had previously been proposed by @lin2023, who developed a 12-item version by removing items with local dependency, high difficulty, or low discrimination. Their parallel analysis supported a unidimensional structure, although this method is known to underestimate the number of factors in multidimensional data [@golino2017exploratory; @markos2024dimensionality]. Consequently, the 12-item IAS was interpreted as reflecting a single underlying factor of interoceptive accuracy.

In contrast, Study 1 applied multiple complementary analytical approaches, including EGA, FA, and HCA. Across datasets and methods, the same tightly coupled item pairs consistently emerged, corresponding to distinct bodily domains such as visceral sensations, musculoskeletal cues, and expulsion phenomena, after several items (e.g., Tickle, Taste, Affective Touch) were excluded due to instability, redundancy, or weak associations.

These findings challenge the assumption that the IAS is unidimensional [@murphy2019; @brand2023]. Whereas @lin2023’s Item Response Theory (IRT)-based approach treated locally dependent items as artefacts to be removed, our results suggest that such dependencies may instead represent theoretically meaningful relationships (e.g., Heart–Breathing). Because IRT assumes unidimensionality [@nguyen2014introduction], this approach may have obscured structured item clusters that capture distinct but interrelated aspects of interoceptive accuracy. Nevertheless, the resulting 14-item IAS from study 1 demonstrated greater interpretability and efficiency, although its psychometric properties require further validation.

<!-- Study 2 findings -->

Study 2 examined the associations between the IAS, its item pairs identified in Study 1, and a range of interoceptive, affective, psychopathological, personality, and related measures. Consistent with previous research, self-reported interoceptive accuracy as assessed by the IAS showed reliable associations with other interoceptive and interoception-related measures [@brand2023; @garfinkel2015knowing; @gaggero2021]. Positive associations were found with interoceptive sensibility, particularly with MAIA facets reflecting noticing and trusting bodily sensations, attention regulation, and awareness of mind–body integration. In contrast, negative associations were observed with measures of interoceptive deficits, such as alexithymia and interoceptive confusion.

Although the relationships between the BPQ–Body Awareness and the IAS, and between the BPQ–Autonomic Reactivity and the IAS, were both significant, they showed opposite directions: the former was positive, whereas the latter was negative. These results align with previous findings reporting a small positive association [@brand2023; @koike2023; @gaggero2021]. However, the literature remains mixed regarding the direction, magnitude, and significance of the relationship between the body awareness subscale and the IAS, with some studies reporting negative associations [@lin2023] and others positive [@brand2023; @gaggero2021; @todd2022] and others quadratic [@campos2021].

Moreover, the Body Awareness subscale has been criticised for its ambiguous instructions, specifically whether respondents are meant to indicate attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of such sensations. As a result, associations with other measures may vary depending on how participants interpret the items [@campos2021; @gabriele2022dissociations; @gaggero2021]. The theoretical focus of the BPQ-BA is also debated: some authors interpret it as assessing awareness of sensations during anxiety [@trevisan2021adaptive], whereas others view it as reflecting general awareness of both neutral and negative sensations [@desmedt2022].

Thus, while we found a positive association between the BPQ–Body Awareness subscale and the IAS, this result should be interpreted with caution. Because we could not assess participants’ understanding of the BPQ items, we recommend that future studies using both the BPQ-BA and the IAS control for participants’ interpretation of the scale to clarify the nature of this relationship.

Robust negative associations also emerged between the IAS and mood-related outcomes, aligning with broader evidence linking anxiety and depression to altered interoceptive processing [e.g., @brand2023; @khalsa2018; @nord2022interoceptive]. However, recent studies have reported null or inconsistent associations between interoceptive abilities and mental health symptomatology, suggesting that these relationships may not generalize across bodily domains or measurement modalities [@banellis2025interoceptive; @jenkinson2024interoception]. For instance, meta-analytic evidence indicates that cardiac interoceptive accuracy tasks are not consistently related to symptoms of depression or anxiety [@desmedt2022does; @jenkinson2024interoception].

Most previous studies of interoceptive accuracy have relied on objective, performance-based tasks, whereas the IAS represents the first subjective measure explicitly designed to assess perceived interoceptive accuracy. However, correlations between IAS scores and objective measures are typically weak or absent [@arsllanova2022; @brand2023; @petzke2024somatic], highlighting a broader debate about how interoceptive accuracy should be conceptualised and measured. Many researchers argue that accuracy can only be assessed objectively through behavioural tasks, and that self-reports of interoceptive ability more appropriately reflect interoceptive sensibility - i.e., the self-perceived tendency to attend to and interpret bodily signals [@khalsa2018; @garfinkel2015knowing].

Others propose that self-report measures should be distinguished according to the specific phenomenon assessed (e.g., attention vs perceived accuracy) to better align with contemporary conceptualisations of interoception [@desmedt2022measures; @murphy2019]. Consistent with this view and the original aim of the validation paper, the IAS is a belief-based measure of one’s interoceptive ability rather than a direct index of accuracy. In other words, the IAS falls within the broader dimension of interoceptive sensibility or interoceptive self-report [@khalsa2018; @suksasilp2022towards], which does not imply a correlation with actual interoceptive performance. Ongoing discussions about terminology and measurement highlight the need for continued refinement to improve the reliability, validity, and conceptual clarity of interoceptive assessments [@desmedt2024conceptualization].

>>>>

<!-- Aims -->
The present mega-analysis provides a comprehensive evaluation of the IAS, examining both its structure (study 1) and its dispositional correlates (study 2). These studies offer novel insights into the conceptual interpretation, and practical applications of the IAS. In this discussion, these findings are contextualized within the broader landscape of interoception research.

<!-- Structural insights from Study 1 -->

Study 1 re-analysed the IAS structure using a combination of traditional psychometric and network-based methods. Across these complementary approaches, seven stable item pairs consistently emerged (i.e., Hungry–Thirsty, Urinate–Defecate, Muscles–Pain, Breathing–Heart, Cough–Sneeze, Wind–Burp, and Bruise–Blood sugar) yielding a refined 14-item structure that demonstrated superior model fit and cross-dataset stability compared to previously proposed unidimensional or bifactor solutions. These findings challenge earlier interpretations of the IAS as reflecting a single latent construct [e.g., @brand2023; @murphy2019] and instead indicate that the scale captures multiple, interrelated but domain-specific clusters of bodily sensations.

A shorter version of the IAS was previously proposed by @lin2023, who developed a 12-item form based on item response theory analyses. Their analysis supported a unidimensional structure, which they interpreted as indexing a single factor of self-reported interoceptive accuracy. In contrast, the present results suggest that apparent local dependencies between items may not represent measurement artefacts but rather theoretically meaningful associations among functionally or phenomenologically coupled sensations (e.g., Heart–Breathing), corresponding to distinct bodily domains such as visceral sensations, musculoskeletal cues, and expulsion phenomena.

Such pairings likely reflect sensations that co-occur physiologically or contextually, such as hunger and thirst, or coughing and sneezing, and tend to be perceived and appraised together. This pattern implies that perceived accuracy within one bodily domain generalizes to related sensations, yielding a clustered rather than continuous organization of interoceptive beliefs.

<!-- there is a paper that states this better but cannot find it yet --> 
Notably, objective interoceptive accuracy tasks across modalities often show weak or absent associations [e.g., @bruni2023interoception; @ferentzi2018interoceptive; @garfinkel2016interoceptive], whereas self-reported interoceptive beliefs appear to cluster more strongly within shared physiological domains [**ref**]. This divergence reinforces the view that subjective interoceptive accuracy may reflect a domain-specific self-belief system rather than a general perceptual ability.

<!-- Dispositional correlates and domain specificity (Study 2) -->

<!-- Interoception and interoception-related -->
Study 2 extended these findings by examining associations between IAS scores from the full scale and the pairings identified in study 1, with a broad set of interoceptive, affective, personality, and belief-related measures.  Consistent with prior research [e.g., @brand2023; @garfinkel2015knowing; @gaggero2021], the IAS showed positive associations with measures of interoceptive sensibility (e.g., MAIA subscales reflecting noticing, attention regulation, and body trust) and negative associations with measures of interoceptive difficulties (e.g., alexithymia and interoceptive confusion).

Although associations between the IAS and the BPQ subscales replicated prior work (e.g., @Brand2023; @Koike2023; @Gaggero2021), interpretation remains uncertain due to conceptual ambiguity in the BPQ–Body Awareness subscale. Prior studies have reported positive, negative, and even quadratic relationships (e.g., @Campos2021; @Lin2023), and some authors have suggested that these mixed findings stem from divergent interpretations of the items—specifically, whether they capture attentional focus on bodily sensations or perceived accuracy in detecting them (e.g., @Gabriele2022Dissociations; @Trevisan2021Adaptive). Thus, although we observed a positive association between the BPQ–Body Awareness and the IAS, this result should be interpreted with caution, and future studies should assess participants’ understanding of BPQ items to clarify the nature of this relationship.

<!-- Mood and psychopathological disorders domain specific -->

Across dispositional constructs, negative associations emerged between self-reported interoceptive accuracy (IAS) and a range of affective and psychopathological traits. Individuals reporting greater symptoms of depression and anxiety also tended to perceive themselves as less accurate in detecting bodily signals. Crucially, these associations were not uniform across bodily domains: the strongest effects consistently involved visceral items (e.g., hunger, thirst), whereas cardioceptive and respiratory items showed weaker or inconsistent patterns. This domain-specific pattern highlights that self-perceived accuracy in internal bodily regulation is more psychologically and clinically relevant than accuracy in reflexive or surface-bodily sensations.

This domain-specificity may help explain inconsistencies in the literature linking interoception and emotional symptomatology. Reviews and meta-analyses indicate that performance-based measures of cardiac interoceptive accuracy, historically the dominant index of interoception, are often unrelated to depression or anxiety [@jenkinson2024interoception; @banellis2025interoceptive; @adams2022association]. In contrast, self-reported interoceptive measures show stronger, albeit sometimes divergent, associations with internalizing traits [e.g., @brand2022; @lin2023; @clemente2024relationship; @murphy2019]. These discrepancies likely reflect the distinction between objective sensitivity and metacognitive self-belief [@khalsa2018; @garfinkel2015knowing; @suksasilp2022towards]. Our results extend this view by demonstrating that even within self-reported interoception, relationships with affective traits are structured along bodily axes, reinforcing the importance of considering domain-specific facets of the IAS.

Beyond mood, similar but weaker patterns emerged across other dispositional traits, again following a visceral-focused gradient. Autistic traits, schizotypal tendencies, and maladaptive personality dimensions generally correlated negatively with IAS scores, primarily for visceral items, whereas cardioceptive and respiratory beliefs contributed little. These findings align with evidence that interoceptive self-beliefs are attenuated in conditions involving social–emotional difficulties or reduced bodily awareness. For example, interoceptive dimensions predict interpersonal-like and disorganized-like schizotypal traits [@pasciucco2025predictive], and autistic individuals report subjective difficulties perceiving internal bodily states such as hunger, thirst, and fatigue, even when objective measures of accuracy are intact [@itoi2023relationship]. More broadly, atypical interoception constitutes a transdiagnostic risk factor for psychopathology, again emphasizing differences across bodily domains [@brewer2023atypical]. Collectively, these results underscore that diminished self-perceived interoceptive accuracy is not a uniform trait but varies systematically by bodily domain, with visceral beliefs showing the clearest psychological relevance.

Taken together, the evidence indicates that the IAS reflects a domain-structured metacognitive construct, complementary to but distinct from objective interoceptive performance. Visceral beliefs—linked to hunger, thirst, or pain—consistently show the strongest relationships with emotional and dispositional traits, whereas cardioceptive and respiratory beliefs contribute little. Recognizing this domain specificity not only clarifies inconsistencies in prior studies emphasizing heartbeat detection but also highlights the importance of assessing a wider range of interoceptive modalities in future research, particularly those related to visceral homeostasis.
